Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-12-10 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits to wgbs_bioinformatics_analysis

This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges. I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol. Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.

@@

Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq. There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis. This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:downstream_analyses}.

@@

Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \citep{Li:2009ka}. However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way. This makes it difficult to develop downstream analysis tools.

@@

Ideally, mapping software assigns the degree of confidence it has that the read is correctly mapped via a mapping quality score (`mapQ`). In theory, reads might be down-weighted in downstream analyses based on the mapping quality score. However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful. `Bismark` \citep{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).

@@

> "Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair. When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment. You can thus think of it as determining "if all the bases from the read were aligned, where would the 5' most base have been aligned". It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair. "Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15" ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).

@@

If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is 'batch aware'. For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with.

@@

I find the first strategy conceptually simpler, and easier to program, particularly when I start dealing with m-tuples, and so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)). In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.

A somewhat subtle point is that M-bias should only be estimated from reads that are actually going to be used for methylation calling. Suppose M-bias is highly correlated with some other quality metric, such as base quality, so that positions with M-bias also have low base quality[^bq_mbias]. If you already intend to ignore all positions with a base quality less than 20 in your methylation calling then it makes sense to also ignore these positions when computing M-bias; otherwise you will overestimate the effect of M-bias and unnecessarily exclude read positions in your methylation calling. Unfortunately, the software I use for estimating M-bias, `bismark_methylation_extractor`, does not allow the user to exclude certain reads or read-positions from its calculations.

@@

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions. It is rather a subjective decision to make. Two problems that I found were in maintaining consistency across samples and in determining where to draw the line when there is no dramatic 'spike' in the M-bias. For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both read 1 and read 2, as well as a big problem at read-position 101 in read 2. What is less clear is where to draw the line on the gradual decay downwards toward the end of read 1 and towards the start of read 2. This motivated me to write a few simple functions to do perform more systematic processing of M-bias results. These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.

@@

To do this, I use a simple normalisation of the read-position methylation levels ($rpml$). Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels (for example, $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read 1_).

The `trim` method identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$. While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types. I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $nrpml_{CpG} < 0.72$ or $nrpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias. It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types. I tend to exclude read-positions that display evidence of M-bias in the methylation type I am working with, which is typically CpG methylation.

@@

Now that we've found read-positions with M-bias, what can we do about it? Typically, read-positions showing evidence of M-bias are excluded when calling methylation events. In fact, a slightly cruder procedure is often used whereby all read-positions preceding or following those with M-bias are excluded. Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method. Generally, this is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads.

However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101. Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.

Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read positions entirely. In the datasets I have analysed, I've had to ignore up to 30 read-positions per read due to M-bias.

One idea is to weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation. The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.

@@

There are several other sources of bias when making methylation calls. Some are what I call technical biases - those driven by the sequencing protocol and processing of the data - while others I call biological biases - those driven by measuring something other than what you thought you were measuring.

@@

For example, many studies of DNA methylation use whole blood as the sample tissue due to the ease with which it can be obtained. However, whole blood contains a mixture of cell types, each of which has a distinct methylation profile. \citet{Houseman:2012km} exploited this fact to estimate the cell composition of whole blood samples, in particular the differences in white blood cell (leukocyte) proportions between different sub-populations (e.g. cases and controls).

@@

The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci at which to call methylation events. It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina ('Lister-style') whole-genome bisulfite-sequencing protocol.

Certain variants, in particular (heterozygous) $C>T$ SNPs, are more difficult to accurately genotype than others. Unfortunately, $C>T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants. Fortunately, it is possible to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand - if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}). Other base substitutions are more readily detected, and insertion and deletion events (indels) may also be called.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Chromosome & Strand & Position & M & U \\ \midrule
chr1       & $+$      & 100      & 7 & 1 \\
chr1       & $-$      & 101      & 5 & 2 \\
chr2       & $+$      & 400      & 0 & 3 \\
chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
\end{tabular}
\caption{Example of output for methylation calling at 1-tuples.}
\label{tab:1-tuples}
\end{table}

@@

\begin{figure}[h!]
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
\label{fig:methtuple_example1}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}

@@

I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples from the EPISCOPE, Lister, Seisenberger and Ziller datasets. To improve the performance I process split each sample by chromosome and process each chromosome in parallel. This is implemented in a helper script ([https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh](https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh)) that makes extensive use of `GNU parallel` \citep{Tange:2011ty}.

>>>>

This chapter and Chapter \ref{chap:wgbs_downstream_analyses} describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges. I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol. Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.

@@

Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq. There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis. This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:wgbs_downstream_analyses}.

@@

Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or its binary equivalent, `BAM` \citep{Li:2009ka}. However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way. This makes it difficult to develop downstream analysis tools.

@@

Ideally, mapping software assigns the degree of confidence it has that the read is correctly mapped via a mapping quality score (`mapQ`). In theory, reads might be down-weighted in downstream analyses based on the mapping quality score. However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful. `Bismark` \citep{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1`, released in April 2014).

@@

> "Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair. When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment. You can thus think of it as determining "if all the bases from the read were aligned, where would the 5' most base have been aligned". It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair. "Best" is defined as the read pair having the highest sum of base qualities as bases with $Q >= 15$" ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).

@@

If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should probably be performed separately for each batch and then combined, or in some manner that is 'batch aware'. For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with.

@@

I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)). In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.

A somewhat subtle point is that M-bias should only be estimated from reads that are actually going to be used for methylation calling. Suppose M-bias is highly correlated with some other quality metric, such as base quality, so that positions with M-bias also have low base quality[^bq_mbias]. If you already intend to ignore all positions with a base quality less than 20 in your methylation calling then it makes sense to also ignore these positions when computing M-bias, otherwise you will overestimate the effect of M-bias and unnecessarily exclude read positions in your methylation calling. Unfortunately, the software I use for estimating M-bias, `bismark_methylation_extractor` \cite{Krueger:2011eb}, does not allow the user to exclude certain reads or read-positions from its calculations.

@@

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions. It is rather a subjective decision to make. Two problems that I found were:

1. Maintaining consistency across samples
2. Determining where to draw the line when there is no dramatic 'spike' in the M-bias.

For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both read 1 and read 2, as well as a big problem at read-position 101 in read 2. What is less clear is where to draw the line on the gradual decay downwards toward the end of read 1 and towards the start of read 2. This motivated me to write a few simple functions to do perform more systematic processing of M-bias results. These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.

@@

To do this, I use a simple normalisation of the read-position methylation levels ($rpml$). Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels, i.e., $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read 1_.

The `trim` method identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$. While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types. I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $rpml_{CpG} < 0.72$ or $rpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias. It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types. I tend to exclude read-positions that display evidence of M-bias in the methylation type I am working with, which is typically CpG methylation.

@@

Now that we've found read-positions with M-bias, what can we do about it? Typically, read-positions showing evidence of M-bias are excluded when calling methylation events. In fact, a slightly cruder procedure is often used whereby the entire ends of reads are removed. For example, for a $100$ bp read with M-bias observed at positions $1-4, 9, 94, 96-100$, then positions $1-9$ and $94-100$ would be ignored when methylation calling.  Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method.

Generally, this strategy is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads. However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101. Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded.

In the datasets I have analysed I've had to ignore up to 30 read-positions per read due to M-bias. Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read positions entirely. One idea is to inversely weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation. The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.

@@

There are several other sources of bias when making methylation calls. Some are what I call technical biases - those driven by the sequencing protocol and processing of the data --- while others I call biological biases - those driven by measuring something other than what you thought you were measuring.

@@

For example, many studies of DNA methylation use whole blood as the sample tissue due to the ease with which it can be obtained. However, whole blood contains a mixture of cell types, each of which has a distinct methylation profile. \citet{Houseman:2012km} exploited this fact to estimate the cell composition of whole blood samples, in particular the differences in white blood cell (leukocyte) proportions between different groups, e.g., cases and controls.

@@

The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci at which to call methylation events. It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina whole-genome bisulfite-sequencing protocol.

Certain variants, in particular heterozygous $C>T$ SNPs, are more difficult to accurately genotype than others. Unfortunately, $C>T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants. Fortunately, it is possible to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand --- if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}). Other base substitutions are more readily detected, and insertion and deletion events (indels) may also be called.

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Chromosome & Strand & Position & M & U \\ \midrule
chr1       & $+$      & 100      & 7 & 1 \\
chr1       & $-$      & 101      & 5 & 2 \\
chr2       & $+$      & 400      & 0 & 3 \\
chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
\end{tabular}
\caption{Example of output for methylation calling at 1-tuples. Most CpGs will have a measurement from both the positive and negative strands, but not always.}
\label{tab:1-tuples}
\end{table}

@@

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
\label{fig:methtuple_example1}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}

@@

I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples from the EPISCOPE, Lister, Seisenberger and Ziller datasets. To improve the performance I  split each sample by chromosome and process each chromosome in parallel. This is implemented in a helper script ([https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh](https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh)) that makes extensive use of `GNU parallel` \citep{Tange:2011ty}.
